---
title:  "Target catch species composition cluster analysis"
output: html_document
---

## Overview

This vignette demonstrates how purse seine sets were classified according to the composition of the three dominant target tuna species:

- Skipjack tuna (SKJ)
- Yellowfin tuna (YFT)
- Bigeye tuna (BET)

Initial data exploration showed these three species made up the overwhelming majority of retained target catch, which is supported by the literature (these are the listed primary target species)

```{r setup}
library(dplyr)
library(ggplot2)
library(ggtern)
library(tidyr)
```

## Prepare target catch composition data

Use the retained catch (MT) of the three main tuna species

Target catch is heavily dominated by SKJ, then YFT, and then BET is much lower

sp_cat_code sp_code retain_tot disc_tot
1 TUN         SKJ       253980.   1143.  
2 TUN         YFT       140624.    526.  
3 TUN         BET         4996.     11.3 
4 TUN         FRI          131.     32.7 
5 NULL        RRU           89.1    35.9

```{r}
# data in wide format with one row per set (set_id)
target_dat <- model_data %>%
  select(set_id, SKJ_MT, YFT_MT, BET_MT)

# cluster on species proportions. Also explored total volumes, but because skipjack is so dominant it takes over - want to capture the known (documented) targeting of bigeye, which is caught in much lower volumes but is higher value 

target_dat <- target_dat %>%  mutate(total_tuna = SKJ_MT, YFT_MT, BET_MT)

# exclude Sets with zero catch of all three species - can't calculatue proportions 
nonzero_sets <- target_dat %>% filter(total_tuna > 0)

# Calculate species proportions:
target_props <- nonzero_sets %>%
  mutate(pSKJ = SKJ_MT / total_tuna,
         pYFT = YFT_MT / total_tuna, 
         pBET = BET_MT / total_tuna
  )
```

## K-means clustering 

```{r}
cluster_dat <- target_props %>% select(pSKJ, pYFT, pBET)

## Use elbow method to decide optimal number of clusters
set.seed(123)

wss <- sapply(
  1:10,
  function(k) {
    kmeans(cluster_dat, centers = k, nstart = 50)$tot.withinss
  })

elbow_df <- data.frame(k = 1:10, wss = wss)

ggplot(elbow_df, aes(k, wss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of clusters (k)", y = "Total within-cluster sum of squares") +
  theme_bw()
# not super obvious but most distinct elbow at k=3

## Fit final clustering solution
set.seed(123)

km <- kmeans(cluster_dat, centers = 3, nstart = 50)

target_props$cluster <- factor(km$cluster)

table(target_props$cluster)
```

# Visualise cluster composition

```{r}
ggtern(
  target_props,
  aes(x = pSKJ, y = pYFT, z = pBET, colour = cluster )) +
  geom_point(alpha = 0.6, size = 1.5) +
  theme_bw() +
  labs(x = "SKJ proportion", y = "YFT proportion", z = "BET proportion",
    colour = "Cluster")
```

# Assign descriptive labels
```{r}
## look at the dominant species composition within each group
target_props %>%
  group_by(cluster) %>%
  summarise(
    mean_SKJ = mean(pSKJ),
    mean_YFT = mean(pYFT),
    mean_BET = mean(pBET)
  )

target_props <- target_props %>%
  mutate(target_cluster = recode(
      cluster,
      "1" = "YFT dominant",
      "2" = "SKJ dominant",
      "3" = "Mixed"))

# Add zero-target category - sets with no catch of SKJ, YFT, or BET
cluster_assignments <- target_props %>% select(set_id, target_cluster)

model_data <- model_data %>%
  left_join(cluster_assignments, by = "set_id") %>%
  mutate(target_cluster = case_when(
      SKJ_MT + YFT_MT + BET_MT == 0 ~ "Zero main targets",
      TRUE ~ target_cluster))

table(model_data$target_cluster)
```

