---
title: "stepwise model selection"

---

# Model selection procedure

```{r setup, include=FALSE}

library(tidyverse)
library(mgcv)
library(MuMIn)
library(broom)

knitr::opts_chunk$set(echo = TRUE)
```

## Example workflow shown for one bycatch species
The same process was repeated independently for all 7
```{r global model}
# data in wide format where 1 row = unique set_ID
mod_global <- gam( bycatch_species ~ # either weight or count
    year + # 2019 - 2023
    s(target_catch_MT) + # combined weight of SKJ, YFT, BET
    s(latitude, longitude) + # coordinates in decimal degrees recorded at start of net setting
    s(month, bs = "cc") + # fishing occurs all months of the year, so use a circular smooth
    s(set_time) + # time set begain (rounded to 1 decimal place, or 10 minute intervals)
    flag + # PNG or Philippines
    eez + # grouped into PNG or other (outside PNG EEZ)
    target_cluster + # k means cluster derived from cluster analysis (4 categories)
    s(set_type, bs = "re") + # 8 categories of associated or unassociated set types
    s(vessel, bs = "re") # unique identifier for the vessel
  data = model_data,
  family = tw(),
  method = "REML"
)
AIC(mod_global)

```

## use dredge to identify candidate supported models

```{r dredge}
options(na.action = "na.fail")

dredge_results <- dredge(mod_global, evaluate = TRUE)

head(dredge_results)
```

## iteratively simplify and compare alternate formations

Where multiple candidate models showed similar support (ΔAIC < 2), we iteratively simplified the model and looked at alternative formats of individual predictors. Model selection considered both AIC and diagnostic performance. 
Where there was a clear winner from the dredge, we also evaluated alternative formats of individual predictors where relevant (e.g. set type as a fixed v random effect)

```{r simplify}

mod_1 <- update(mod_global, .~. - flag - eez)
AIC(mod_global, mod_1)

```

## check altnerate distriubtions

bycatch observations were veryzero-inflated and overdispersed - compare Tweedie and negative binomial distributions 
Tweedie models were consistently better, and worked for all response variables (silky sharks as weights, all other models as counts)

```{r compare distributions}
mod_tw <- mod_1
mod_nb <- update(mod_tw, family = nb())
AIC(mod_tw, mod_nb)
```

## Fixed effect vessel factor
Test vessel as fixed instead of random effect
Try penalizing vessel term (more relevant for the rarer species)
```{r vessel fixed effect}

# first just normal fixed effect
mod_2 <- update(mod_1, .~. - s(vessel, bs = "re") + vessel)

# then set penalty on vessel factor
NoCoefs <- length(unique(model_data$vessel))-1 
pp <- list(vessel = list(rank = NoCoefs, diag=diag(NoCoefs)))
# # set sum to zero contrasts
contrasts(model_data$vessel) <- contr.sum(NoCoefs + 1)

# refit instead of using update to make sure the contrast coding works
mod_3 <- gam(bycatch_species ~
               s(x) +
               y +
               vessel,
  data = model_data,
  family = tw(),
  method = "REML",
  paraPen = pp)

```

## Model diagnostics

For each model:
1. look at model summaries and AIC values.
2. check basis dimensions and residual patterns using `gam.check()`.
3. check concurvity among model terms using `concurvity()`.
4. look at DHARMa simulated residuals and test for uniformity, dispersion, outliers
5. inspect fitted smooth functions using `gratia::draw()`.

```{r}
mods <- list(
  mobulids = mobulids.best,
  silky = silky.best,
  turtles = turtles.best,
  whaleshark = whaleshark.best,
  fkw = fkw.best,
  dolphins = dolphins.best,
  baleens = baleens.best
)

sim_res <- simulateResiduals(
  fittedModel = mod,
  n = 1000
)

gam.check(mod) # adjust ks where needed

concurvity(mod, full = TRUE)

testUniformity(sim_res)
testDispersion(sim_res)
testOutliers(sim_res)

gratia::draw(mod)

# save outputs
mapply(
  run_gam_diagnostics,
  mod = mods,
  name = names(mods),
  MoreArgs = list(outdir = outdir)
)
```
